---
title: "CAG-170-analysis"
output: html_document
date: "2026-07-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(networkD3)
library(htmlwidgets)
library(scales)
library(ggrepel)
library(ape)
library(phangorn)
library(vegan)
library(tidyverse)
library(patchwork)
library(webshot2)
library(RColorBrewer)
library(forcats)
library(tibble)
library(writexl)
library(panstripe)
```

## Load data
# Complete (1,490 genomes) and quality-filtered (562) metadata tables

```{r load-complete}
metadata_complete <- read_tsv("metadata_complete.tsv", show_col_types = FALSE)

metadata_complete %>% count(Species, sort = TRUE)
metadata_complete %>% count(Continent, sort = TRUE)
metadata_complete %>% count(Animal, sort = TRUE)
```

```{r load-filtered}
metadata_filtered <- read_tsv("metadata_filtered.tsv", show_col_types = FALSE)

metadata_filtered %>% count(Species, sort = TRUE)
metadata_filtered %>% count(Continent, sort = TRUE)
metadata_filtered %>% count(Animal, sort = TRUE)
```

## Species grouping (top 8 + "Other")

```{r species-grouping}
# Top 8 species in each dataset
top8_species_complete <- metadata_complete %>%
  filter(!is.na(Species), !Species %in% c("Unknown", "NA", "-", "")) %>%
  count(Species, sort = TRUE) %>%
  slice_head(n = 8) %>%
  pull(Species)

top8_species_filtered <- metadata_filtered %>%
  filter(!is.na(Species), !Species %in% c("Unknown", "NA", "-", "")) %>%
  count(Species, sort = TRUE) %>%
  slice_head(n = 8) %>%
  pull(Species)

# "Other"
classify_species_plot <- function(species_vec, top8) {
  species_vec <- trimws(as.character(species_vec))
  case_when(
    is.na(species_vec)                           ~ "Other",
    species_vec %in% c("", "-", "NA", "Unknown") ~ "Other",
    species_vec %in% top8                        ~ species_vec,
    TRUE                                         ~ "Other"
  )
}

metadata_complete <- metadata_complete %>%
  mutate(Species_plot = classify_species_plot(Species, top8_species_complete))

metadata_filtered <- metadata_filtered %>%
  mutate(Species_plot = classify_species_plot(Species, top8_species_filtered))

species_union <- union(top8_species_complete, top8_species_filtered)
species_colours <- brewer.pal(length(species_union), "Paired")

palette_species <- c(
  setNames(species_colours, species_union),
  "Other" = "#DBEDF4"
)
```

```{r species-grouping-check}
metadata_complete %>% count(Species_plot, sort = TRUE)
metadata_filtered %>% count(Species_plot, sort = TRUE)
unique(metadata_filtered$Species_plot)
```

## Palettes

```{r palettes}
COL_HUMAN   <- "#D3D3D3"
COL_UNKNOWN <- "#DBEDF4"

# Host palette
palette_animal <- c(
  "Human"      = COL_HUMAN,
  "Pig"        = brewer.pal(8, "Set2")[1],
  "Mouse"      = brewer.pal(8, "Set2")[2],
  "Chimpanzee" = brewer.pal(8, "Set2")[3],
  "Cattle"     = brewer.pal(8, "Set2")[4],
  "Wild boar"  = brewer.pal(8, "Set2")[5],
  "Trout"      = brewer.pal(8, "Set2")[6],
  "Sheep"      = brewer.pal(8, "Set2")[7],
  "Water deer" = brewer.pal(8, "Set2")[8],
  "Unknown"    = COL_UNKNOWN
)

# Continent palette
palette_continent <- c(
  setNames(
    brewer.pal(6, "Accent"),
    c("Europe", "Asia", "North America", "Oceania", "Africa", "South America")
  ),
  "Unknown" = COL_UNKNOWN
)

# Genome quality palette
palette_quality <- c("High" = "#99c945", "Medium" = "#f0e442")

# Binary host palette
palette_host_binary <- c(
  "Human"   = COL_HUMAN,
  "Animal"  = brewer.pal(8, "Set2")[2],
  "Unknown" = COL_UNKNOWN
)

# Enrichment direction palette
palette_direction <- c("Enriched" = "#4477AA", "Depleted" = "#CC6666")

# Database source palette
palette_source <- setNames(
  brewer.pal(4, "Pastel1"),
  c("MGnify", "GTDB", "Pig MAG catalogue", "Isolate")
)

FONT_FAMILY <- "Arial"
```

## Consistent plotting theme

```{r plotting-theme}
# Consistent plotting for A4
FS_TITLE         <- 15
FS_AXIS_TITLE    <- 10
FS_AXIS_TEXT     <- 9
FS_AXIS_TEXT_ROT <- 8
FS_BAR_LABEL     <- 3.2 
FS_LEGEND_TITLE  <- 9.5
FS_LEGEND        <- 9
FS_COG_AXIS_TEXT <- 7 
FS_VOLCANO_AXIS  <- 9

theme_clean <- theme_minimal(base_size = 9) +
  theme(
    plot.title = element_text(face = "bold", size = FS_TITLE, colour = "black",
                              hjust = 0.5, margin = margin(b = 6)),
    axis.title = element_text(size = FS_AXIS_TITLE, face = "bold", colour = "black"),
    axis.text  = element_text(size = FS_AXIS_TEXT, colour = "black"),
    legend.title = element_text(size = FS_LEGEND_TITLE, colour = "black"),
    legend.text  = element_text(size = FS_LEGEND, colour = "black"),
    panel.grid = element_blank(),
    legend.position = "right",
    plot.margin = margin(t = 5, r = 5, b = 5, l = 5)
  )

theme_clean_rotated_x <- theme_clean +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1,
                               size = FS_AXIS_TEXT_ROT, colour = "black")
  )

update_geom_defaults("col", list(colour = NA))
update_geom_defaults("bar", list(colour = NA))

save_fig <- function(filename_base, plot, width = 17, height = 10, dpi = 300) {
  ggsave(paste0(filename_base, ".png"), plot = plot,
         width = width, height = height, units = "cm", dpi = dpi, bg = "white")
  ggsave(paste0(filename_base, ".pdf"), plot = plot,
         width = width, height = height, units = "cm", device = cairo_pdf)
}
```

## Metadata summary bar charts

```{r panels-complete-source}
source_counts <- metadata_complete %>%
  filter(!is.na(Source)) %>%
  count(Source, sort = TRUE)

fig_source_counts <- ggplot(source_counts, aes(x = reorder(Source, -n), y = n, fill = Source)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.4, size = FS_BAR_LABEL, colour = "black") +
  scale_fill_brewer(palette = "Pastel1") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Genomes by database source", x = "Source", y = "Number of genomes") +
  theme_clean_rotated_x +
  theme(legend.position = "none")
```

```{r panels-complete-continent}
continent_counts <- metadata_complete %>%
  mutate(Continent = ifelse(is.na(Continent), "Unknown", Continent)) %>%
  count(Continent, sort = TRUE) %>%
  mutate(Continent = fct_reorder(Continent, -n),
         Continent = fct_relevel(Continent, "Unknown", after = Inf))

fig_continent_counts <- ggplot(continent_counts, aes(x = Continent, y = n, fill = Continent)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.4, size = FS_BAR_LABEL, colour = "black") +
  scale_fill_manual(values = palette_continent, na.value = "#DBEDF4") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Genomes by continent", x = "Continent", y = "Number of genomes") +
  theme_clean_rotated_x +
  theme(legend.position = "none")
```

```{r panels-complete-host}
animal_counts <- metadata_complete %>%
  mutate(Animal = ifelse(is.na(Animal), "Unknown", Animal)) %>%
  count(Animal, sort = TRUE) %>%
  mutate(Animal = fct_reorder(Animal, -n),
         Animal = fct_relevel(Animal, "Unknown", after = Inf))

fig_animal_counts <- ggplot(animal_counts, aes(x = Animal, y = n, fill = Animal)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.4, size = FS_BAR_LABEL, colour = "black") +
  scale_fill_manual(values = palette_animal, na.value = "#DBEDF4") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Genomes by host", x = "Host", y = "Number of genomes") +
  theme_clean_rotated_x +
  theme(legend.position = "none")
```

```{r panels-complete-species}
species_counts <- metadata_complete %>%
  filter(!is.na(Species_plot), Species_plot != "Unknown") %>%
  count(Species_plot, sort = TRUE) %>%
  mutate(Species_plot = fct_reorder(Species_plot, -n),
         Species_plot = fct_relevel(Species_plot, "Other", after = Inf))

fig_species_counts <- ggplot(species_counts, aes(x = Species_plot, y = n, fill = Species_plot)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.4, size = FS_BAR_LABEL, colour = "black") +
  scale_fill_manual(values = palette_species, na.value = "#DBEDF4") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Genomes by species", x = "Species", y = "Number of genomes") +
  theme_clean_rotated_x +
  theme(legend.position = "none")
```

# Filtered dataset panels

```{r panels-filtered-source}
# add "Source" to filtered dataset from complete dataset
if (!"Source" %in% colnames(metadata_filtered)) {
  metadata_filtered <- metadata_filtered %>%
    left_join(metadata_complete %>% select(Genome, Source), by = "Genome")
}

source_counts_filtered <- metadata_filtered %>%
  filter(!is.na(Source)) %>%
  count(Source, sort = TRUE)

fig_source_counts_filtered <- ggplot(source_counts_filtered, aes(x = reorder(Source, -n), y = n, fill = Source)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.4, size = FS_BAR_LABEL, colour = "black") +
  scale_fill_brewer(palette = "Pastel1") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Genomes by database source (dereplicated dataset, n=562)",
       x = "Source", y = "Number of genomes") +
  theme_clean_rotated_x +
  theme(legend.position = "none")
```

```{r panels-filtered-host}
animal_counts_filtered <- metadata_filtered %>%
  mutate(Animal = ifelse(is.na(Animal), "Unknown", Animal)) %>%
  count(Animal, sort = TRUE) %>%
  mutate(Animal = fct_reorder(Animal, -n),
         Animal = fct_relevel(Animal, "Unknown", after = Inf))

fig_animal_counts_filtered <- ggplot(animal_counts_filtered, aes(x = Animal, y = n, fill = Animal)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.4, size = FS_BAR_LABEL, colour = "black") +
  scale_fill_manual(values = palette_animal, na.value = "#DBEDF4") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Genomes by host (dereplicated dataset, n=562)",
       x = "Host", y = "Number of genomes") +
  theme_clean_rotated_x +
  theme(legend.position = "none")
```

```{r panels-filtered-continent}
continent_counts_filtered <- metadata_filtered %>%
  mutate(Continent = ifelse(is.na(Continent), "Unknown", Continent)) %>%
  count(Continent, sort = TRUE) %>%
  mutate(Continent = fct_reorder(Continent, -n),
         Continent = fct_relevel(Continent, "Unknown", after = Inf))

fig_continent_counts_filtered <- ggplot(continent_counts_filtered, aes(x = Continent, y = n, fill = Continent)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.4, size = FS_BAR_LABEL, colour = "black") +
  scale_fill_manual(values = palette_continent, na.value = "#DBEDF4") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Genomes by continent (dereplicated dataset, n=562)",
       x = "Continent", y = "Number of genomes") +
  theme_clean_rotated_x +
  theme(legend.position = "none")
```

```{r panels-filtered-species}
species_counts_filtered <- metadata_filtered %>%
  filter(!is.na(Species_plot), Species_plot != "Unknown") %>%
  count(Species_plot, sort = TRUE) %>%
  mutate(Species_plot = fct_reorder(Species_plot, -n),
         Species_plot = fct_relevel(Species_plot, "Other", after = Inf))

fig_species_counts_filtered <- ggplot(species_counts_filtered, aes(x = Species_plot, y = n, fill = Species_plot)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.4, size = FS_BAR_LABEL, colour = "black") +
  scale_fill_manual(values = palette_species, na.value = "#DBEDF4") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Genomes by species (dereplicated dataset, n=562)",
       x = "Species", y = "Number of genomes") +
  theme_clean_rotated_x +
  theme(legend.position = "none")
```

# Combined 2×2 grids

```{r grid-complete}
fig_genomes_complete <- (
  (fig_source_counts    + labs(title = NULL)) +
  (fig_animal_counts    + labs(title = NULL)) +
  (fig_continent_counts + labs(title = NULL)) +
  (fig_species_counts   + labs(title = NULL))
) +
  plot_layout(ncol = 2, nrow = 2, guides = "collect") +
  plot_annotation(
    title = "Complete dataset genomes (n=1,490)",
    theme = theme(plot.title = element_text(face = "bold", size = FS_TITLE, hjust = 0.5))
  )

fig_genomes_complete
save_fig("fig_genomes_complete_2x2", fig_genomes_complete, width = 17, height = 15)
```

```{r grid-filtered}
fig_genomes_filtered <- (
  (fig_source_counts_filtered    + labs(title = NULL, x = NULL)) +
  (fig_animal_counts_filtered    + labs(title = NULL, x = NULL)) +
  (fig_continent_counts_filtered + labs(title = NULL, x = NULL)) +
  (fig_species_counts_filtered   + labs(title = NULL, x = NULL))
) +
  plot_layout(ncol = 2, nrow = 2, guides = "collect", axis_titles = "collect") +
  plot_annotation(
    title = "Filtered dataset genomes (n=562)",
    theme = theme(plot.title = element_text(face = "bold", size = FS_TITLE, hjust = 0.5))
  )

fig_genomes_filtered
save_fig("fig_genomes_filtered_2x2", fig_genomes_filtered, width = 17, height = 15)
```

## Sankey plot

```{r sankey}
sankey_data <- metadata_complete %>%
  mutate(Source = recode(Source, "Pig_catalogue" = "Pig MAG catalogue")) %>%
  filter(!is.na(Animal), !is.na(Continent), !is.na(Source))

links1 <- sankey_data %>%
  dplyr::count(Source, Continent, name = "value") %>%
  dplyr::rename(source = Source, target = Continent)

links2 <- sankey_data %>%
  dplyr::count(Continent, Animal, name = "value") %>%
  dplyr::rename(source = Continent, target = Animal)

links <- bind_rows(links1, links2)

nodes <- data.frame(name = unique(c(links$source, links$target))) %>%
  mutate(
    group = name,
    colour = case_when(
      name %in% names(palette_source)    ~ unname(palette_source[name]),
      name %in% names(palette_continent) ~ unname(palette_continent[name]),
      name %in% names(palette_animal)    ~ unname(palette_animal[name]),
      TRUE ~ "#e3e3e3"
    )
  )

links$IDsource <- match(links$source, nodes$name) - 1
links$IDtarget <- match(links$target, nodes$name) - 1

colour_domain <- paste0('["', paste(nodes$name, collapse = '","'), '"]')
colour_range  <- paste0('["', paste(nodes$colour, collapse = '","'), '"]')
ColourScal <- paste0('d3.scaleOrdinal().domain(', colour_domain, ').range(', colour_range, ')')

SANKEY_WIDTH  <- 1800
SANKEY_HEIGHT <- 1340

sankey_plot <- sankeyNetwork(
  Links = links, Nodes = nodes,
  Source = "IDsource", Target = "IDtarget", Value = "value",
  NodeID = "name", NodeGroup = "group",
  sinksRight = FALSE, colourScale = ColourScal,
  nodeWidth = 12, nodePadding = 28,
  fontSize = FS_LEGEND * 4, fontFamily = "Arial",
  width = SANKEY_WIDTH, height = SANKEY_HEIGHT
) %>%
  htmlwidgets::onRender(
    "function(el) {
       d3.select(el).selectAll('.link')
         .style('stroke', '#999999')
         .style('stroke-opacity', 0.25);
     }"
  )

sankey_plot

sankey_html_path <- file.path(getwd(), "metadata_sankey_plot.html")
saveWidget(sankey_plot, sankey_html_path, selfcontained = TRUE)

webshot2::webshot(sankey_html_path, "metadata_sankey_plot.png",
                  vwidth = SANKEY_WIDTH, vheight = SANKEY_HEIGHT, delay = 0.5)
```

## Assembly quality
# median, N50 range, contig count, and >500 contig count

```{r qc-summary}
qc_summary_table <- bind_rows(
  metadata_complete %>%
    filter(!is.na(N_contigs), !is.na(N50)) %>%
    summarise(
      Dataset        = "Complete (n=1,490)",
      N50_median     = median(N50),
      N50_min        = min(N50),
      N50_max        = max(N50),
      contig_median  = median(N_contigs),
      contig_min     = min(N_contigs),
      contig_max     = max(N_contigs),
      n_over_500     = sum(N_contigs > 500)
    ),
  metadata_filtered %>%
    filter(!is.na(N_contigs), !is.na(N50)) %>%
    summarise(
      Dataset        = "Filtered (n=562)",
      N50_median     = median(N50),
      N50_min        = min(N50),
      N50_max        = max(N50),
      contig_median  = median(N_contigs),
      contig_min     = min(N_contigs),
      contig_max     = max(N_contigs),
      n_over_500     = sum(N_contigs > 500)
    )
)

print(qc_summary_table)
```

```{r qc-boxplots}
qc_data <- metadata_complete %>%
  select(Genome, N_contigs, N50) %>%
  filter(!is.na(N_contigs), !is.na(N50))

p_contigs_box <- ggplot(qc_data, aes(y = N_contigs)) +
  geom_boxplot(fill = "#A6CEE3", colour = "black") +
  scale_y_log10() +
  labs(title = "Contig count distribution", y = "Number of contigs (log10)", x = NULL) +
  theme_clean +
  theme(axis.text.x = element_blank())

p_n50_box <- ggplot(qc_data, aes(y = N50)) +
  geom_boxplot(fill = "#A6CEE3", colour = "black") +
  scale_y_log10() +
  labs(title = "N50 distribution", y = "N50 (log10)", x = NULL) +
  theme_clean +
  theme(axis.text.x = element_blank())

p_contigs_box + p_n50_box
```

```{r qc-histograms}
p_contigs_hist <- ggplot(qc_data, aes(x = N_contigs)) +
  geom_histogram(bins = 100, fill = "#A6CEE3", colour = "black", linewidth = 0.2) +
  geom_vline(xintercept = 500, linetype = "dashed", colour = "#B2182B", linewidth = 0.7) +
  labs(title = "Contig count histogram", x = "Number of contigs", y = "Number of genomes") +
  theme_clean

p_n50_hist <- ggplot(qc_data, aes(x = N50)) +
  geom_histogram(bins = 100, fill = "#A6CEE3", colour = "black", linewidth = 0.2) +
  scale_x_log10() +
  labs(title = "N50 histogram (log10)", x = "N50 (log10)", y = "Number of genomes") +
  theme_clean

p_contigs_hist + p_n50_hist

ggsave("p_contigs_hist.pdf", p_contigs_hist, width = 7, height = 5)
ggsave("p_n50_hist.pdf", p_n50_hist, width = 7, height = 5)
```

## Binary host variable

```{r binary-host}
host_levels <- c("Human", "Animal", "Unknown")

metadata_complete <- metadata_complete %>%
  mutate(Host_binary = factor(case_when(
    Animal == "Human" ~ "Human",
    is.na(Animal) | trimws(as.character(Animal)) %in% c("", "-", "NA", "Unknown") ~ "Unknown",
    TRUE ~ "Animal"
  ), levels = host_levels))

metadata_filtered <- metadata_filtered %>%
  mutate(Host_binary = factor(case_when(
    Animal == "Human" ~ "Human",
    is.na(Animal) | trimws(as.character(Animal)) %in% c("", "-", "NA", "Unknown") ~ "Unknown",
    TRUE ~ "Animal"
  ), levels = host_levels))
```

## Species composition by host

```{r species-composition-function}
make_species_composition <- function(
    data, fill_var, palette, legend_name, title_topic, filename,
    stack_order = NULL,
    plot_mode = "both",
    show_n = TRUE,
    include_unknown = FALSE,
    species_order = NULL
) {

  if (include_unknown) {
    data <- data %>%
      mutate({{ fill_var }} := ifelse(is.na({{ fill_var }}), "Unknown", {{ fill_var }}))
  }

  if (!is.null(stack_order) && include_unknown) {
    stack_order <- c(stack_order, "Unknown")
  }

  plot_data <- data %>%
    filter(!is.na(Species_plot), Species_plot != "Unknown", !is.na({{ fill_var }})) %>%
    count(Species_plot, {{ fill_var }}) %>%
    group_by(Species_plot) %>%
    mutate(total = sum(n), proportion = n / total) %>%
    ungroup()

  if (!is.null(species_order)) {
    plot_data <- plot_data %>%
      mutate(Species_plot = factor(Species_plot, levels = species_order))
  } else {
    plot_data <- plot_data %>%
      mutate(Species_plot = reorder(Species_plot, -total),
             Species_plot = forcats::fct_relevel(Species_plot, "Other", after = Inf))
  }

  if (!is.null(stack_order)) {
    plot_data <- plot_data %>%
      mutate({{ fill_var }} := factor({{ fill_var }}, levels = stack_order))
  }

  totals <- plot_data %>% distinct(Species_plot, total)

  p_count <- ggplot(plot_data, aes(x = Species_plot, y = n, fill = {{ fill_var }})) +
    geom_col(width = 0.8, position = position_stack(reverse = TRUE)) +
    geom_text(data = totals, aes(x = Species_plot, y = total, label = total),
              inherit.aes = FALSE, vjust = -0.4, size = FS_BAR_LABEL,
              colour = "black", family = FONT_FAMILY) +
    scale_fill_manual(values = palette, na.value = "#DBEDF4", name = legend_name) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
    labs(title = paste("Genome counts per CAG-170 species by", title_topic),
         x = NULL, y = "Number of genomes") +
    theme_clean_rotated_x

  p_prop <- ggplot(plot_data, aes(x = Species_plot, y = proportion, fill = {{ fill_var }})) +
    geom_col(width = 0.8, position = position_stack(reverse = TRUE)) +
    scale_fill_manual(values = palette, na.value = "#DBEDF4", name = legend_name) +
    scale_y_continuous(breaks = c(0.25, 0.50, 0.75, 1.00),
                       labels = scales::percent_format(accuracy = 1),
                       limits = c(0, 1), expand = expansion(mult = c(0, 0.12))) +
    labs(title = paste("Proportion per CAG-170 species by", title_topic),
         x = "Species", y = "Proportion") +
    theme_clean_rotated_x

  if (show_n) {
    p_prop <- p_prop +
      geom_text(data = totals, aes(x = Species_plot, y = 1, label = paste0("n = ", total)),
                inherit.aes = FALSE, vjust = -0.5, size = FS_BAR_LABEL,
                colour = "black", family = FONT_FAMILY)
  }

  if (plot_mode == "count") {
    fig <- p_count
  } else if (plot_mode == "proportion") {
    fig <- p_prop
  } else {
    fig <- (p_count / p_prop + plot_layout(guides = "collect")) &
      theme(legend.position = "right")
  }

  filename_base <- tools::file_path_sans_ext(filename)
  save_fig(filename_base, fig, width = 17, height = ifelse(plot_mode == "both", 16, 6))

  fig
}
```

```{r species-composition}
species_order_filtered <- levels(species_counts_filtered$Species_plot)

fig_species_host_prop <- make_species_composition(
  metadata_filtered, Host_binary, palette_host_binary, "Host", "host",
  "fig_species_host_prop.pdf",
  stack_order = c("Human", "Animal"),
  plot_mode = "proportion",
  species_order = species_order_filtered
)

fig_species_animal_prop <- make_species_composition(
  metadata_filtered, Animal, palette_animal, "Host", "host",
  "fig_species_animal_prop.pdf",
  stack_order = c("Human", "Pig", "Mouse", "Chimpanzee", "Cattle",
                  "Wild boar", "Trout", "Sheep", "Water deer"),
  include_unknown = FALSE,
  plot_mode = "proportion",
  species_order = species_order_filtered
)

# Stacked vertically with x-axis species labels only on the bottom panel
fig_species_all_proportions <- (
  (fig_species_animal_prop + labs(title = NULL, x = NULL) +
     theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())) /
  (fig_species_host_prop   + labs(title = NULL, x = NULL))
) +
  plot_layout(guides = "collect", axis_titles = "collect") +
  plot_annotation(
    title = stringr::str_wrap("CAG-170 species composition by host", width = 45),
    theme = theme(plot.title = element_text(face = "bold", family = FONT_FAMILY,
                                            size = 12, hjust = 0.5))
  ) &
  theme(legend.position = "right")

fig_species_all_proportions
save_fig("fig_species_all_proportions", fig_species_all_proportions, width = 17, height = 10)
```

## Phylogenetic trees (midpoint rooting)
# Load, midpoint-root, and save for iTOL visualisation
```{r trees-root}
tree_complete <- read.tree("tree_complete.nwk")
tree_filtered  <- read.tree("tree_filtered.nwk")

tree_complete_rooted <- midpoint(tree_complete)
tree_filtered_rooted  <- midpoint(tree_filtered)

write.tree(tree_complete_rooted, "tree_1490_midpoint_rooted.nwk")
write.tree(tree_filtered_rooted,  "tree_562_midpoint_rooted.nwk")
```

```{r trees-continents}
# Recode blank/NA continent to "Unknown" for iTOL colour strips
metadata_complete <- metadata_complete %>%
  mutate(Continent = ifelse(is.na(Continent) | Continent %in% c("", "-", "NA"),
                            "Unknown", Continent))

metadata_filtered <- metadata_filtered %>%
  mutate(Continent = ifelse(is.na(Continent) | Continent %in% c("", "-", "NA"),
                            "Unknown", Continent))
```

## iTOL annotation exports

```{r itol-colorstrip-function}
write_itol_colorstrip <- function(data, id_col, value_col, output_file,
                                  dataset_label, colour_map,
                                  strip_width = 120, margin = 20, label_size_factor = 9) {
  data_out <- data %>%
    mutate(
      value  = .data[[value_col]],
      colour = colour_map[value],
      colour = ifelse(is.na(colour), "#CCCCCC", colour)
    ) %>%
    transmute(id = .data[[id_col]], colour, label = value)

  # Legend list the categories that are present in the dataset
  present_values <- names(colour_map)[names(colour_map) %in% unique(data_out$label)]
  legend_colors <- unname(colour_map[present_values])
  legend_shapes <- rep(1, length(present_values))

  writeLines(c(
    "DATASET_COLORSTRIP", "SEPARATOR TAB",
    paste0("DATASET_LABEL\t", dataset_label),
    "COLOR\t#000000", "SHOW_STRIP_LABELS\t0",
    paste0("STRIP_WIDTH\t", strip_width),
    paste0("MARGIN\t", margin),
    "SHOW_LABELS\t1",
    paste0("SIZE_FACTOR\t", label_size_factor),
    paste0("LEGEND_TITLE\t", dataset_label),
    paste0("LEGEND_SHAPES\t", paste(legend_shapes, collapse = "\t")),
    paste0("LEGEND_COLORS\t", paste(legend_colors, collapse = "\t")),
    paste0("LEGEND_LABELS\t", paste(present_values, collapse = "\t")),
    "", "DATA",
    paste(data_out$id, data_out$colour, data_out$label, sep = "\t")
  ), output_file)

  message("Written: ", output_file)
}
```

# Flagging genomes with more than 500 contigs and failed GUNC in complete dataset
```{r itol-qc-markers}
qc_flags <- metadata_complete %>%
  mutate(
    Tree_ID      = str_replace(Genome, "\\.fa$", ""),
    high_contigs = ifelse(N_contigs > 500, 1, -1),
    gunc_fail    = ifelse(GUNC_pass == "no", 1, -1)
  ) %>%
  filter(high_contigs == 1 | gunc_fail == 1)

writeLines(c(
  "DATASET_BINARY", "SEPARATOR TAB",
  "DATASET_LABEL\tGenome QC markers", "COLOR\t#000000",
  "FIELD_SHAPES\t3\t3",
  "FIELD_LABELS\tN_contigs > 500\tFailed GUNC",
  "FIELD_COLORS\t#E9B200\t#461D7C",
  "MAXIMUM_SIZE\t15",
  "SHOW_LABELS\t0", "MARGIN\t20", "HEIGHT_FACTOR\t8",
  "SYMBOL_SPACING\t-25",
  "HORIZONTAL_GRID\t0",
  "VERTICAL_GRID\t0",
  "LEGEND_TITLE\tQC flags",
  "LEGEND_SHAPES\t3\t3",
  "LEGEND_COLORS\t#E9B200\t#461D7C",
  "LEGEND_LABELS\tN_contigs > 500\tFailed GUNC",
  "DATA",
  paste(qc_flags$Tree_ID, qc_flags$high_contigs, qc_flags$gunc_fail, sep = "\t")
), "itol_qc_markers.txt")
message("Written: itol_qc_markers.txt")
```

```{r itol-clades-1490}
clade_members <- list(
  "Clade I" = c(
    "S93310B_bin_30", "F5381_bin_58", "MGYG000303586", "GCA_016297855.1",
    "MGYG000301861", "GCA_945916265.1", "LW1-s151260_bin_10", "LW2-s127530_bin_14",
    "LW1-s151250_bin_39", "GCA_022770025.1", "GCA_004555655.1", "LW2-s126572_bin_26",
    "LW1-s150971_bin_18", "W4930_bin_59", "MGYG000301454", "GCA_945876125.1"
  ),
  "Clade II" = c(
    "MGYG000301360", "GCA_945874875.1", "GCA_022779855.1", "MGYG000301662",
    "GCA_945910055.1", "GCA_034087285.1", "GCA_028725385.1", "GCA_029012045.1",
    "MGYG000302872", "GCA_946012855.1"
  ),
  "Clade III" = c(
    "W4950_bin_7", "GCA_004556885.1", "GCA_022781885.1", "GCA_017938825.1",
    "GCA_034087745.1", "GCA_028722805.1", "GCA_034117885.1", "GCA_004559455.1",
    "GCA_004554725.1", "GCA_022774265.1", "MGYG000301729", "GCA_945911325.1",
    "MGYG000301121", "GCA_945870135.1", "MGYG000302830", "GCA_946007635.1",
    "GCA_028725405.1", "GCA_022771045.1", "MGYG000299607", "GCA_947072225.1",
    "MGYG000303269", "GCA_946409655.1", "GCA_004554295.1", "GCA_022774875.1",
    "MGYG000301158", "MGYG000300246", "GCA_945870995.1", "GCA_945827745.1"
  )
)

clade_members <- lapply(clade_members, stringr::str_replace_all, " ", "_")
colours <- c("Clade I" = "#FDC9B4", "Clade II" = "#FDC9B4", "Clade III" = "#FDC9B4")

data_lines <- sapply(names(clade_members), function(clade) {
  tips <- clade_members[[clade]]
  col  <- colours[[clade]]
  paste(dplyr::first(tips), dplyr::last(tips), col, "", col, "solid", 1,
        clade, "#000000", 1, "bold", sep = ",")
})

writeLines(c(
  "DATASET_RANGE", "SEPARATOR COMMA",
  "DATASET_LABEL,Animal-associated clades",
  "COLOR,#FC8D62", "RANGE_TYPE,box", "RANGE_COVER,clade",
  "COVER_LABELS,0", "COVER_DATASETS,0", "FIT_LABELS,0",
  "SHOW_LABELS,1", "LABEL_POSITION,top-right",
  "LABELS_VERTICAL,0", "STRAIGHT_LABELS,1",
  "LEGEND_TITLE,Animal-associated clades",
  paste0("LEGEND_SHAPES,", paste(rep(1, 3), collapse = ",")),
  paste0("LEGEND_COLORS,", paste(colours, collapse = ",")),
  paste0("LEGEND_LABELS,", paste(names(colours), collapse = ",")),
  "DATA", data_lines
), "itol_animal_clades.txt")
```

# Range boxes for animal-associated clades (genome IDs identified from the visualised trees)
```{r itol-clades-562}
clade_members_filtered <- list(
  "Clade I" = c(
    "GCA_016297855.1", "GCA_945916265.1", "GCA_022770025.1", "GCA_004555655.1",
    "LW2-s127530_bin_14", "LW1-s151260_bin_10", "LW2-s126572_bin_26"
  ),
  "Clade II" = c(
    "GCA_029012045.1", "GCA_034087285.1", "GCA_028725385.1",
    "GCA_945874875.1", "GCA_945910055.1"
  ),
  "Clade III" = c(
    "W4950_bin_7", "GCA_004556885.1", "GCA_028722805.1", "GCA_022781885.1",
    "GCA_004554725.1", "GCA_022774265.1", "GCA_034117885.1", "GCA_022774875.1",
    "GCA_028725405.1", "GCA_004554295.1", "GCA_029226425.1", "GCA_022771045.1",
    "GCA_037248705.1", "MGYG000144480", "MGYG000034682"
  )
)

clade_members_filtered <- lapply(clade_members_filtered, stringr::str_replace_all, " ", "_")
colours <- c("Clade I" = "#FDC9B4", "Clade II" = "#FDC9B4", "Clade III" = "#FDC9B4")

data_lines <- sapply(names(clade_members_filtered), function(clade) {
  tips <- clade_members_filtered[[clade]]
  col  <- colours[[clade]]
  paste(dplyr::first(tips), dplyr::last(tips), col, "", col, "solid", 1,
        clade, "#000000", 1, "bold", sep = ",")
})

writeLines(c(
  "DATASET_RANGE", "SEPARATOR COMMA",
  "DATASET_LABEL,Animal-associated clades",
  "COLOR,#FC8D62", "RANGE_TYPE,box", "RANGE_COVER,clade",
  "COVER_LABELS,0", "COVER_DATASETS,0", "FIT_LABELS,0",
  "SHOW_LABELS,1", "LABEL_POSITION,top-right",
  "LABELS_VERTICAL,0", "STRAIGHT_LABELS,1",
  "LEGEND_TITLE,Animal-associated clades",
  paste0("LEGEND_SHAPES,", paste(rep(1, 3), collapse = ",")),
  paste0("LEGEND_COLORS,", paste(colours, collapse = ",")),
  paste0("LEGEND_LABELS,", paste(names(colours), collapse = ",")),
  "DATA", data_lines
), "itol_animal_clades_562.txt")
```

```{r itol-colorstrips-1490}
metadata_itol <- metadata_complete %>%
  mutate(Species_itol = ifelse(Species_plot == "Unknown", "Other", Species_plot))

write_itol_colorstrip(metadata_itol, "Genome", "Species_itol",
                      "itol_species.txt", "Species", palette_species, label_size_factor = 12)
write_itol_colorstrip(metadata_itol, "Genome", "Host_binary",
                      "itol_host_binary.txt", "Host", palette_host_binary, label_size_factor = 12)
write_itol_colorstrip(metadata_itol, "Genome", "Continent",
                      "itol_continent.txt", "Continent", palette_continent, label_size_factor = 12)
write_itol_colorstrip(metadata_itol, "Genome", "CheckM_quality",
                      "itol_genome_quality.txt", "Genome quality", palette_quality, label_size_factor = 12)
```

```{r itol-colorstrips-562}
metadata_filtered_itol <- metadata_filtered %>%
  mutate(Species_itol = Species_plot)

write_itol_colorstrip(metadata_filtered_itol, "Genome", "Host_binary",
                      "itol_host_binary_562.txt", "Host", palette_host_binary, label_size_factor = 9)
write_itol_colorstrip(metadata_filtered_itol, "Genome", "Continent",
                      "itol_continent_562.txt", "Continent", palette_continent, label_size_factor = 9)
write_itol_colorstrip(metadata_filtered_itol, "Genome", "Species_itol",
                      "itol_species_562.txt", "Species", palette_species, label_size_factor = 9)
```

## Per-continent phylogenetic diversity (complete dataset)

```{r diversity-distances}
continents_to_test <- c("Europe", "Asia", "North America", "Africa", "Oceania", "South America")

diversity_distances_complete <- tibble()

for (cont in continents_to_test) {
  genomes_in_cont <- metadata_complete %>% filter(Continent == cont) %>% pull(Genome)
  tips_present    <- intersect(genomes_in_cont, tree_complete$tip.label)

  subtree     <- keep.tip(tree_complete, tips_present)
  dist_matrix <- cophenetic.phylo(subtree)

  cont_distances <- tibble(
    Continent         = cont,
    n_genomes         = length(tips_present),
    pairwise_distance = as.vector(as.dist(dist_matrix))
  )

  diversity_distances_complete <- bind_rows(diversity_distances_complete, cont_distances)
}
```

```{r diversity-boxplot}
fig_continent_diversity_complete <- ggplot(
  diversity_distances_complete,
  aes(x = reorder(Continent, pairwise_distance, median),
      y = pairwise_distance, fill = Continent)
) +
  geom_boxplot(colour = "grey30", linewidth = 0.3, outlier.size = 0.6, outlier.alpha = 0.5) +
  scale_fill_manual(values = palette_continent, na.value = "#DBEDF4") +
  labs(x = NULL, y = "Pairwise cophenetic distance") +
  theme_clean +
  theme(
    legend.position    = "none",
    axis.text.x        = element_text(size = 15, face = "bold", angle = 30, hjust = 1),
    axis.text.y        = element_text(size = 14, face = "bold"),
    axis.title.y       = element_text(size = 16, face = "bold"),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank()
  )

fig_continent_diversity_complete
save_fig("fig_continent_diversity1490", fig_continent_diversity_complete, width = 18, height = 12)
```

## PCoA of accessory genome content
# Normalise genome IDs (for the ones starting with "GCA_")

```{r pangenome-fix-function}
fix_isolate_id <- function(mat) {
  colnames(mat) <- ifelse(
    grepl("^GCA_", colnames(mat)),
    sub("^(GCA_[0-9]+\\.[0-9]+).*", "\\1", colnames(mat)),
    colnames(mat)
  )
  colnames(mat)[colnames(mat) == "GCA_916988195.3"] <- "GCA_916988195.3_POC01_20211130"
  mat
}
```

# Build the accessory-genome Jaccard distance, run PCoA, and k-means clustering
```{r pcoa-compute}
pan <- read.delim("gene_presence_absence_filtered.Rtab", header = TRUE,
                  row.names = 1, check.names = FALSE)
pan_matrix <- fix_isolate_id(as.matrix(pan))

gene_prevalence <- rowSums(pan_matrix) / ncol(pan_matrix)
pan_filtered    <- pan_matrix[gene_prevalence >= 0.15 & gene_prevalence <= 0.95, ]
pan_t           <- t(pan_filtered)

jaccard_dist <- vegdist(pan_t, method = "jaccard")
pcoa         <- cmdscale(jaccard_dist, k = 2, eig = TRUE)

pcoa_df <- data.frame(
  Genome = rownames(pan_t), PC1 = pcoa$points[, 1], PC2 = pcoa$points[, 2]
) %>%
  left_join(
    metadata_filtered %>%
      select(Genome, Animal, Continent, Species_plot, Host_binary),
    by = "Genome"
  )

for (k in 2:4) {
  set.seed(123)
  pcoa_df[[paste0("k", k)]] <- as.character(kmeans(pcoa_df[, c("PC1", "PC2")], centers = k)$cluster)
}
pcoa_df$Cluster <- pcoa_df$k3
```

# Run elbow method to identify optimal number of clusters
```{r pcoa-elbow}
wss <- sapply(1:10, function(k) {
  set.seed(123)
  kmeans(pcoa_df[, c("PC1", "PC2")], centers = k, nstart = 1)$tot.withinss
})
elbow_df <- data.frame(k = 1:10, wss = wss)

fig_elbow <- ggplot(elbow_df, aes(x = k, y = wss)) +
  geom_line() +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Elbow plot for k-means clustering",
       x = "Number of clusters (k)", y = "Within-cluster sum of squares") +
  theme_clean

fig_elbow
save_fig("fig_elbow", fig_elbow, width = 17, height = 9)
```

```{r pcoa-kmeans}
palette_kmeans <- setNames(brewer.pal(3, "Set1"), c("1", "2", "3"))

pcoa_kmeans <- ggplot(pcoa_df, aes(x = PC1, y = PC2, colour = Cluster)) +
  geom_point(size = 2) +
  scale_colour_manual(values = palette_kmeans, name = "Cluster") +
  labs(title = "PCoA of accessory genome content (k=3)") +
  theme_clean

pcoa_kmeans
ggsave("pcoa.png", pcoa_kmeans, width = 7, height = 5)
```

```{r pcoa-host}
pcoa_df_host_ordered <- pcoa_df %>% arrange(Host_binary == "Animal")

pcoa_host <- ggplot(pcoa_df_host_ordered, aes(x = PC1, y = PC2, colour = Host_binary)) +
  geom_point(size = 2) +
  scale_colour_manual(values = palette_host_binary, na.value = "#DBEDF4", name = "Host") +
  labs(title = "PCoA of accessory genome content, coloured by host") +
  theme_clean

pcoa_host
ggsave("pcoa_host.png", pcoa_host, width = 7, height = 5)
```

```{r pcoa-continent}
pcoa_continent <- ggplot(pcoa_df, aes(x = PC1, y = PC2, colour = Continent)) +
  geom_point(size = 2) +
  scale_colour_manual(values = palette_continent, na.value = "#DBEDF4", name = "Continent") +
  labs(title = "PCoA of accessory genome content, coloured by continent") +
  theme_clean

pcoa_continent
ggsave("pcoa_continent.png", pcoa_continent, width = 7, height = 5)
```

```{r pcoa-species}
pcoa_species <- ggplot(pcoa_df, aes(x = PC1, y = PC2, colour = Species_plot)) +
  geom_point(size = 2) +
  scale_colour_manual(values = palette_species, na.value = "#DBEDF4", name = "Species") +
  labs(title = "PCoA of accessory genome content, coloured by species") +
  theme_clean

pcoa_species
ggsave("pcoa_species.png", pcoa_species, width = 7, height = 5)
```

```{r pcoa-combined}
pcoa_kmeans <- pcoa_kmeans +
  labs(title = NULL) +
  guides(colour = guide_legend(nrow = 1, byrow = TRUE)) +
  theme(legend.text = element_text(size = 8),
        legend.title = element_text(size = 9, face = "bold"),
        legend.key.width = grid::unit(0.3, "cm"),
        legend.key.height = grid::unit(0.3, "cm"))

pcoa_host <- pcoa_host +
  labs(title = NULL) +
  guides(colour = guide_legend(nrow = 1, byrow = TRUE)) +
  theme(legend.text = element_text(size = 8),
        legend.title = element_text(size = 9, face = "bold"),
        legend.key.width = grid::unit(0.3, "cm"),
        legend.key.height = grid::unit(0.3, "cm"))

pcoa_continent <- pcoa_continent +
  labs(title = NULL) +
  guides(colour = guide_legend(nrow = 3, byrow = TRUE)) +
  theme(legend.text = element_text(size = 6.5),
        legend.title = element_text(size = 8, face = "bold"),
        legend.key.width = grid::unit(0.22, "cm"),
        legend.key.height = grid::unit(0.22, "cm"))

pcoa_species <- pcoa_species +
  labs(title = NULL) +
  guides(colour = guide_legend(nrow = 5, byrow = TRUE)) +
  theme(legend.text = element_text(size = 6.5),
        legend.title = element_text(size = 8, face = "bold"),
        legend.key.width = grid::unit(0.22, "cm"),
        legend.key.height = grid::unit(0.22, "cm"))

fig_pcoa_combined <- (pcoa_kmeans + pcoa_host + pcoa_continent + pcoa_species) +
  plot_layout(ncol = 2, nrow = 2, guides = "keep",
              axes = "collect", axis_titles = "collect") +
  plot_annotation(
    title = "Principal coordinate analysis of accessory-genome composition",
    theme = theme(plot.title = element_text(size = 15, face = "bold", colour = "black",
                                            hjust = 0.5, margin = margin(b = 8)))
  ) &
  theme(panel.border = element_rect(colour = "grey50", fill = NA, linewidth = 0.6),
        legend.position = "bottom", legend.direction = "horizontal",
        legend.spacing.x = grid::unit(0.08, "cm"),
        plot.margin = margin(4, 4, 4, 4))

fig_pcoa_combined

ggsave("fig_pcoa_combined.png", fig_pcoa_combined,
       width = 17, height = 15, units = "cm", dpi = 300, bg = "white")
```

## PERMANOVA
# Using cophenetic distances derived from phylogenetic trees; each variable is tested individually.

```{r permanova-phylo}
run_permanova_phylo <- function(variable, metadata_obj, tree_obj, dataset_label) {

  common_ids  <- intersect(tree_obj$tip.label, metadata_obj$Genome)
  tree_pruned <- keep.tip(tree_obj, common_ids)

  meta_sub <- metadata_obj %>%
    filter(Genome %in% tree_pruned$tip.label,
           !is.na(.data[[variable]]),
           !.data[[variable]] %in% c("Unknown", "NA", "-")) %>%
    distinct(Genome, .keep_all = TRUE) %>%
    arrange(match(Genome, tree_pruned$tip.label))

  phylo_dist <- cophenetic.phylo(tree_pruned)
  dist_sub   <- as.dist(phylo_dist[meta_sub$Genome, meta_sub$Genome])
  f          <- as.formula(paste("dist_sub ~", variable))

  set.seed(123)
  result <- adonis2(f, data = meta_sub, permutations = 999)

  tibble(
    Variable   = variable,
    R2_percent = result$R2[1] * 100,
    p_value    = result$`Pr(>F)`[1],
    Dataset    = dataset_label
  )
}

vars <- c("Continent", "Country", "Animal", "Species")

permanova_complete <- bind_rows(
  lapply(vars, run_permanova_phylo,
         metadata_obj = metadata_complete, tree_obj = tree_complete,
         dataset_label = "Full dataset (n=1,490)")
)

permanova_filtered <- bind_rows(
  lapply(vars, run_permanova_phylo,
         metadata_obj = metadata_filtered, tree_obj = tree_filtered,
         dataset_label = "Dereplicated subset (n=562)")
)

permanova_complete %>% print(n = Inf)
permanova_filtered  %>% print(n = Inf)
```

```{r permanova-phylo-1490-plot}
permanova_phylo_plot <- permanova_complete %>%
  filter(Variable %in% c("Animal", "Continent")) %>%
  mutate(
    Variable = recode(Variable, "Animal" = "Host", "Continent" = "Continent"),
    Variable = reorder(Variable, R2_percent),
    sig_label = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE            ~ "ns"
    ),
    bar_label = paste0(round(R2_percent, 1), "% ", sig_label)
  )

fig_permanova_phylo <- ggplot(permanova_phylo_plot, aes(x = R2_percent, y = Variable)) +
  geom_col(fill = "#999999", width = 0.6, colour = "grey30", linewidth = 0.3) +
  geom_text(aes(label = bar_label), hjust = -0.1, size = 6.5,
            colour = "black", fontface = "bold") +
  labs(x = expression("Variance explained (R"^2*", %)"), y = NULL) +
  theme_clean +
  theme(panel.grid.major.y = element_blank(),
        axis.text  = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 17, face = "bold")) +
  expand_limits(x = max(permanova_phylo_plot$R2_percent) * 1.3)

fig_permanova_phylo
save_fig("fig_permanova_phylo", fig_permanova_phylo, width = 16, height = 4)
```

```{r permanova-phylo-562-plot}
permanova_filtered_plot <- permanova_filtered %>%
  filter(Variable %in% c("Animal", "Continent")) %>%
  mutate(
    Variable = recode(Variable, "Animal" = "Host", "Continent" = "Continent"),
    Variable = reorder(Variable, R2_percent),
    sig_label = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE            ~ "ns"
    ),
    bar_label = paste0(round(R2_percent, 1), "% ", sig_label)
  )

fig_permanova_filtered <- ggplot(permanova_filtered_plot, aes(x = R2_percent, y = Variable)) +
  geom_col(fill = "#0050A0", width = 0.6, colour = "grey30", linewidth = 0.3) +
  geom_text(aes(label = bar_label), hjust = -0.1, size = 6.5,
            colour = "black", fontface = "bold") +
  labs(x = expression("Variance explained (R"^2*", %)"), y = NULL) +
  theme_clean +
  theme(panel.grid.major.y = element_blank(),
        axis.text  = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 17, face = "bold")) +
  expand_limits(x = max(permanova_filtered_plot$R2_percent) * 1.3)

fig_permanova_filtered
save_fig("fig_permanova_562", fig_permanova_filtered, width = 16, height = 4)
```

# Using Jaccard distances for accessory genome; each variable is tested individually.
```{r permanova-jaccard-compute}
run_permanova_jaccard <- function(variable) {
  dist_ids <- rownames(as.matrix(jaccard_dist))

  meta_sub <- metadata_filtered %>%
    filter(Genome %in% dist_ids,
           !is.na(.data[[variable]]),
           !.data[[variable]] %in% c("Unknown", "NA", "-")) %>%
    arrange(match(Genome, dist_ids))

  dist_sub <- as.dist(as.matrix(jaccard_dist)[meta_sub$Genome, meta_sub$Genome])
  f        <- as.formula(paste("dist_sub ~", variable))

  set.seed(123)
  result <- adonis2(f, data = meta_sub, permutations = 999)

  tibble(
    Variable   = variable,
    R2_percent = result$R2[1] * 100,
    p_value    = result$`Pr(>F)`[1]
  )
}

permanova_jaccard <- bind_rows(
  lapply(c("Species", "Continent", "Animal"),
         run_permanova_jaccard)
) %>%
  mutate(
    Variable = recode(Variable,
      "Animal"      = "Host"),
    Variable  = reorder(Variable, R2_percent)
  )

cat("=== PERMANOVA: Jaccard distance (accessory genome, n=562) ===\n")
print(permanova_jaccard)
```


## Pangenome accumulation and panstripe (visualisation)
```{r pangenome-load}
pa_genus <- pan_matrix

# Species-level presence/absence matrix
pa_sp002404795_raw <- read.delim("gene_presence_absence_sp002404795.Rtab",
                                 header = TRUE, row.names = 1, check.names = FALSE)

pa_sp002404795_raw <- as.matrix(fix_isolate_id(pa_sp002404795_raw))
```

```{r pangenome-subset}
sp002404795_genomes <- metadata_filtered %>% filter(Species == "sp002404795") %>% pull(Genome)

tree_sp002404795_full <- keep.tip(tree_filtered, intersect(sp002404795_genomes, tree_filtered$tip.label))

common_sp002404795 <- intersect(tree_sp002404795_full$tip.label, colnames(pa_sp002404795_raw))

tree_sp002404795 <- keep.tip(tree_sp002404795_full, common_sp002404795)

pa_sp002404795 <- pa_sp002404795_raw[, common_sp002404795, drop = FALSE]
```

```{r accumulation-curve-function}
PANGENOME_COLOUR  <- "#009E73"
COREGENOME_COLOUR <- "#D55E00"

make_acc_curve <- function(pa_obj, subtitle_label, n_iterations = 100) {
  set.seed(123)
  n_genomes <- nrow(pa_obj)
  acc_results <- vector("list", n_iterations)

  for (i in seq_len(n_iterations)) {
    order_i    <- sample(n_genomes)
    pa_ordered <- pa_obj[order_i, , drop = FALSE]
    cum_mat <- apply(pa_ordered, 2, cumsum)
    if (!is.matrix(cum_mat)) cum_mat <- matrix(cum_mat, nrow = n_genomes)
    pan_size  <- rowSums(cum_mat >= 1)
    core_size <- sapply(seq_len(n_genomes), function(k) sum(cum_mat[k, ] == k))
    acc_results[[i]] <- data.frame(n_genomes = seq_len(n_genomes),
                                   pan_size = pan_size, core_size = core_size, iteration = i)
  }

  acc_summary <- bind_rows(acc_results) %>%
    group_by(n_genomes) %>%
    summarise(
      pan_mean = mean(pan_size), pan_q025 = quantile(pan_size, 0.025), pan_q975 = quantile(pan_size, 0.975),
      core_mean = mean(core_size), core_q025 = quantile(core_size, 0.025), core_q975 = quantile(core_size, 0.975),
      .groups = "drop"
    )

  ggplot(acc_summary, aes(x = n_genomes)) +
    geom_ribbon(aes(ymin = pan_q025, ymax = pan_q975), fill = PANGENOME_COLOUR, alpha = 0.15) +
    geom_line(aes(y = pan_mean, colour = "Pan-genome"), linewidth = 0.8) +
    geom_ribbon(aes(ymin = core_q025, ymax = core_q975), fill = COREGENOME_COLOUR, alpha = 0.15) +
    geom_line(aes(y = core_mean, colour = "Core genome"), linewidth = 0.8) +
    scale_colour_manual(values = c("Pan-genome" = PANGENOME_COLOUR,
                                   "Core genome" = COREGENOME_COLOUR), name = NULL) +
    labs(x = "Number of genomes", y = "Number of gene clusters", subtitle = subtitle_label) +
    theme_clean +
    theme(legend.position = "top")
}
```

```{r accumulation-curves}
fig_acc_all         <- make_acc_curve(t(pan_matrix),      "Genus level (n=562)")
fig_acc_sp002404795 <- make_acc_curve(t(pa_sp002404795),  "sp002404795")

fig_acc_all
fig_acc_sp002404795
```

```{r panstripe-check}
stopifnot(all(colnames(pa_genus)       %in% tree_filtered$tip.label))
stopifnot(all(colnames(pa_sp002404795) %in% tree_sp002404795$tip.label))
```

```{r pangenome-genus-fig}
fig_u_all <- plot_u(t(pa_genus), tree_filtered)
fig_u_all$layers[[1]]$aes_params$fill <- PANGENOME_COLOUR
fig_u_all <- fig_u_all + theme_clean + labs(title = NULL)

fig_pangenome_genus <- (
  (fig_acc_all + labs(subtitle = NULL)) + fig_u_all
) +
  plot_layout(nrow = 1, guides = "collect") +
  plot_annotation(
    title = "Genus-level pangenome",
    theme = theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))
  ) &
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", hjust = 0.5))

fig_pangenome_genus
save_fig("fig_pangenome_genus", fig_pangenome_genus, width = 17, height = 8)
```

```{r pangenome-species-fig}
fig_u_sp002404795 <- plot_u(t(pa_sp002404795), tree_sp002404795)
fig_u_sp002404795$layers[[1]]$aes_params$fill <- PANGENOME_COLOUR
fig_u_sp002404795 <- fig_u_sp002404795 + theme_clean + labs(title = NULL)

fig_pangenome_species <- (
  (fig_acc_sp002404795 + labs(subtitle = NULL)) + fig_u_sp002404795
) +
  plot_layout(nrow = 1, guides = "collect") +
  plot_annotation(
    title = "Species-level pangenome",
    theme = theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))
  ) &
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", hjust = 0.5))

fig_pangenome_species
save_fig("fig_pangenome_species", fig_pangenome_species, width = 17, height = 8)
```

## GWAS: human vs non-human

```{r gwas-files}
pyseer_files <- tribble(
  ~label,               ~file,                                   ~direction_label, ~title,
  "human_nocovar",      "pyseer_human_results.tsv",              "human",          "Human vs non-human",
  "human_covar",        "pyseer_human_covar_results.tsv",        "human",          "Human vs non-human (species-adjusted)",
  "sp002404795_human",  "pyseer_sp002404795_human_results.tsv",  "human",          "Human vs non-human (species-level GWAS)"
)

read_and_fdr <- function(file) {
  read_tsv(file, show_col_types = FALSE) %>%
    filter(!is.na(`lrt-pvalue`), `lrt-pvalue` > 0) %>%
    mutate(p_fdr = p.adjust(`lrt-pvalue`, method = "fdr"))
}

pyseer_data <- lapply(pyseer_files$file, read_and_fdr)
names(pyseer_data) <- pyseer_files$label
```


```{r gwas-annotation-sources}
read_eggnog <- function(file) {
  raw <- read_tsv(file, comment = "##", show_col_types = FALSE)
  raw %>%
    rename(variant = `#query`, preferred = Preferred_name, cog = COG_category, eggnog_desc = Description) %>%
    select(variant, preferred, cog, eggnog_desc) %>%
    mutate(
      preferred   = ifelse(preferred   == "-" | is.na(preferred),   NA_character_, preferred),
      cog         = ifelse(cog         == "-" | is.na(cog),         NA_character_, cog),
      eggnog_desc = ifelse(eggnog_desc == "-" | is.na(eggnog_desc), NA_character_, eggnog_desc)
    )
}

eggnog_filtered <- read_eggnog("eggnog.emapper.annotations_filtered")
eggnog_sp002404795 <- read_eggnog("eggnog.emapper.annotations_sp002404795")

eggnog_lookup <- list(
  human_nocovar     = eggnog_filtered,
  human_covar       = eggnog_filtered,
  sp002404795_human = eggnog_sp002404795
)
```

```{r cog-labels}
cog_labels <- c(
  # INFORMATION STORAGE AND PROCESSING
  J = "Translation, ribosomal structure and biogenesis",
  A = "RNA processing and modification",
  K = "Transcription",
  L = "Replication, recombination and repair",
  B = "Chromatin structure and dynamics",
  # CELLULAR PROCESSES AND SIGNALING
  D = "Cell cycle control, cell division, chromosome partitioning",
  Y = "Nuclear structure",
  V = "Defense mechanisms",
  T = "Signal transduction mechanisms",
  M = "Cell wall/membrane/envelope biogenesis",
  N = "Cell motility",
  Z = "Cytoskeleton",
  W = "Extracellular structures",
  U = "Intracellular trafficking, secretion, and vesicular transport",
  O = "Posttranslational modification, protein turnover, chaperones",
  # METABOLISM
  C = "Energy production and conversion",
  G = "Carbohydrate transport and metabolism",
  E = "Amino acid transport and metabolism",
  F = "Nucleotide transport and metabolism",
  H = "Coenzyme transport and metabolism",
  I = "Lipid transport and metabolism",
  P = "Inorganic ion transport and metabolism",
  Q = "Secondary metabolites biosynthesis, transport and catabolism",
  # POORLY CHARACTERIZED
  R = "General function prediction only",
  S = "Function unknown"
)
```

```{r gwas-annotate-function}
annotate_sig_hits_eggnog <- function(label) {
  direction_label <- pyseer_files$direction_label[pyseer_files$label == label]

  pyseer_data[[label]] %>%
    filter(p_fdr < 0.05) %>%
    left_join(eggnog_lookup[[label]], by = "variant") %>%
    mutate(
      has_eggnog_name = !is.na(preferred) &
        !str_detect(preferred, regex("hypothetical", ignore_case = TRUE)),
      function_desc = ifelse(
        !is.na(eggnog_desc) &
          !str_detect(eggnog_desc, regex("hypothetical|uncharacterized", ignore_case = TRUE)),
        eggnog_desc, NA_character_
      ),
      cog_category = case_when(
        is.na(cog) | cog == "-"   ~ "No COG match",
        str_sub(cog, 1, 1) == "S" ~ "S - Function unknown",
        TRUE ~ paste0(str_sub(cog, 1, 1), " - ", cog_labels[str_sub(cog, 1, 1)])
      ),
      direction = as.character(ifelse(beta > 0,
                        paste("Human-associated", direction_label),
                        paste("Non-human-associated", direction_label)))
    ) %>%
    filter(has_eggnog_name) %>%
    arrange(`lrt-pvalue`)
}
```

```{r gwas-top20-tables}
make_top20_eggnog_table <- function(label) {
  annotate_sig_hits_eggnog(label) %>%
    slice_head(n = 20) %>%
    transmute(
      `Gene`     = preferred,
      `Function` = function_desc,
      `COG`      = cog_category,
      `Beta`     = round(beta, 3),
      `p (FDR)`  = formatC(p_fdr, format = "e", digits = 2)
    )
}

hits_human_covar_eggnog <- make_top20_eggnog_table("human_covar")
hits_sp002404795_eggnog         <- make_top20_eggnog_table("sp002404795_human")

hits_human_covar_eggnog
hits_sp002404795_eggnog
```

```{r gwas-volcano-function}
make_volcano_top20_eggnog <- function(label, top_n = 20, plot_title = NULL) {

  meta          <- pyseer_files %>% filter(label == !!label)
  deplete_label <- "Non-human-associated"
  enrich_label  <- "Human-associated"
  title_to_use  <- if (is.null(plot_title)) meta$title else plot_title

  df <- pyseer_data[[label]] %>%
    left_join(eggnog_lookup[[label]], by = "variant") %>%
    mutate(
      log10p          = -log10(`lrt-pvalue`),
      significant     = p_fdr < 0.05,
      has_eggnog_name = !is.na(preferred) &
        !str_detect(preferred, regex("hypothetical", ignore_case = TRUE)),
      display_cat = case_when(
        significant & has_eggnog_name & beta > 0 ~ enrich_label,
        significant & has_eggnog_name & beta < 0 ~ deplete_label,
        TRUE ~ "Not significant"
      )
    )

  n_sig       <- sum(df$significant)
  n_annotated <- sum(df$significant & df$has_eggnog_name)

  cat(meta$label, "- significant:", n_sig, "| eggNOG-annotated:", n_annotated, "\n")
  if (n_annotated == 0) { message("No eggNOG-annotated hits for: ", meta$label); return(NULL) }

  top_hits <- df %>% filter(significant & has_eggnog_name) %>%
    slice_min(`lrt-pvalue`, n = top_n) %>% pull(variant)
  df <- df %>% mutate(label_text = ifelse(variant %in% top_hits, preferred, NA))

  threshold_y <- df %>% filter(significant) %>%
    summarise(t = max(`lrt-pvalue`)) %>% pull(t)

  ggplot(df, aes(x = beta, y = log10p, colour = display_cat)) +
    geom_point(size  = ifelse(df$display_cat != "Not significant", 3.0, 0.4),
               alpha = ifelse(df$display_cat != "Not significant", 1.0, 0.25)) +
    geom_hline(yintercept = -log10(threshold_y), linetype = "dashed",
               colour = "grey40", linewidth = 0.5) +
    geom_vline(xintercept = 0, linetype = "dotted", colour = "grey40", linewidth = 0.4) +
    geom_text_repel(aes(label = label_text), size = 4.5, max.overlaps = 30,
                    box.padding = 0.5, segment.size = 0.3, segment.colour = "grey50",
                    fontface = "italic", na.rm = TRUE, min.segment.length = 0.2,
                    show.legend = FALSE) +
    scale_colour_manual(
      values = setNames(
        c(palette_direction[["Enriched"]], palette_direction[["Depleted"]], "#DDDDDD"),
        c(enrich_label, deplete_label, "Not significant")
      ),
      breaks = c(deplete_label, enrich_label, "Not significant")
    ) +
    guides(colour = guide_legend(override.aes = list(size = 4, alpha = 1))) +
    labs(title = title_to_use, x = "Effect size (beta)",
         y = expression(-log[10](p)), colour = NULL) +
    theme_clean +
    theme(legend.position = "top",
          axis.text  = element_text(size = FS_VOLCANO_AXIS),
          axis.title = element_text(size = FS_VOLCANO_AXIS, face = "bold"))
}
```

```{r gwas-volcano-plots}
p_human_covar_eggnog <- make_volcano_top20_eggnog("human_covar",
                                                  plot_title = "Genus-level host GWAS")
p_human_covar_eggnog

p_sp002404795_eggnog <- make_volcano_top20_eggnog("sp002404795_human",
                                          plot_title = "Species-level host GWAS")
p_sp002404795_eggnog

save_fig("fig_v_human_covar_eggnog", p_human_covar_eggnog)
save_fig("fig_v_sp002404795_eggnog", p_sp002404795_eggnog)
```

```{r gwas-cog-function}
make_cog_directional <- function(label, plot_title = NULL, aspect_ratio = 1) {

  meta          <- pyseer_files %>% filter(label == !!label)
  deplete_label <- "Non-human-associated"
  enrich_label  <- "Human-associated"
  title_to_use  <- if (is.null(plot_title)) meta$title else plot_title

  df_all <- pyseer_data[[label]] %>% filter(p_fdr < 0.05)
  n_enrich_total  <- sum(df_all$beta > 0)
  n_deplete_total <- sum(df_all$beta < 0)

  if (nrow(df_all) == 0) {
    message("No significant hits for: ", meta$label, " - skipping COG plot")
    return(NULL)
  }

  df <- df_all %>%
    left_join(eggnog_lookup[[label]], by = "variant") %>%
    filter(!is.na(cog), cog != "-", str_sub(cog, 1, 1) != "S") %>%
    mutate(
      direction   = ifelse(beta > 0, enrich_label, deplete_label),
      cog_primary = str_sub(cog, 1, 1),
      cog_label   = cog_labels[cog_primary]
    )

  cog_summary <- df %>%
    group_by(cog_label, direction) %>%
    summarise(n = n(), .groups = "drop") %>%
    mutate(
      prop = case_when(
        direction == enrich_label  ~ n / n_enrich_total,
        direction == deplete_label ~ n / n_deplete_total
      ),
      prop_signed = ifelse(direction == deplete_label, -prop, prop)
    )

  cog_order <- cog_summary %>% filter(direction == enrich_label) %>%
    arrange(prop) %>% pull(cog_label)
  all_cats  <- unique(cog_summary$cog_label)

  complete_df <- expand.grid(cog_label = all_cats,
                             direction = c(enrich_label, deplete_label),
                             stringsAsFactors = FALSE) %>%
    left_join(cog_summary, by = c("cog_label", "direction")) %>%
    mutate(
      n = replace_na(n, 0), prop = replace_na(prop, 0), prop_signed = replace_na(prop_signed, 0),
      cog_label = factor(cog_label, levels = unique(c(cog_order, setdiff(all_cats, cog_order))))
    )

  x_limit <- max(abs(complete_df$prop_signed), na.rm = TRUE) * 1.3

  ggplot(complete_df, aes(x = prop_signed, y = cog_label, fill = direction)) +
    geom_col(width = 0.7, colour = "grey30", linewidth = 0.2) +
    geom_vline(xintercept = 0, colour = "grey30", linewidth = 0.5) +
    geom_text(
      data = complete_df %>% filter(n > 0),
      aes(label = n, x = prop_signed + ifelse(prop_signed >= 0, x_limit * 0.03, -x_limit * 0.03),
          hjust = ifelse(prop_signed >= 0, 0, 1)),
      size = FS_BAR_LABEL, colour = "black"
    ) +
    scale_fill_manual(
      values = setNames(
        c(palette_direction[["Enriched"]], palette_direction[["Depleted"]]),
        c(enrich_label, deplete_label)
      ),
      breaks = c(deplete_label, enrich_label)
    ) +
    scale_x_continuous(labels = function(x) paste0(abs(round(x * 100, 0)), "%"),
                       limits = c(-x_limit, x_limit)) +
    labs(title = title_to_use, x = "Proportion of all significant hits", y = NULL, fill = NULL) +
    theme_clean +
    theme(legend.position = "top",
          panel.grid.major.y = element_blank(),
          axis.text.y = element_text(size = FS_COG_AXIS_TEXT + 2, lineheight = 1))
}
```

```{r gwas-cog-plots}
p_cog_genus   <- make_cog_directional("human_covar",
                                      plot_title = "COG: genus-level host GWAS")
p_cog_species <- make_cog_directional("sp002404795_human",
                                      plot_title = "COG: species-level host GWAS")

p_cog_genus
p_cog_species

save_fig("fig_cog_human_covar", p_cog_genus, height = 11)
save_fig("fig_cog_sp002404795_human", p_cog_species, height = 7)
```

```{r gwas-cog-table-function}
make_cog_directional_table <- function(label) {

  enrich_label  <- "Enriched"
  deplete_label <- "Depleted"

  df_all <- pyseer_data[[label]] %>% filter(p_fdr < 0.05)
  n_enrich_total  <- sum(df_all$beta > 0)
  n_deplete_total <- sum(df_all$beta < 0)

  if (nrow(df_all) == 0) return(tibble())

  df <- df_all %>%
    left_join(eggnog_lookup[[label]], by = "variant") %>%
    filter(!is.na(cog), cog != "-", str_sub(cog, 1, 1) != "S") %>%
    mutate(
      direction   = ifelse(beta > 0, enrich_label, deplete_label),
      cog_primary = str_sub(cog, 1, 1),
      cog_label   = cog_labels[cog_primary]
    )

  df %>%
    group_by(cog_label, direction) %>%
    summarise(n_genes = n(), .groups = "drop") %>%
    mutate(
      proportion = case_when(
        direction == enrich_label  ~ n_genes / n_enrich_total,
        direction == deplete_label ~ n_genes / n_deplete_total
      ),
      `proportion (%)` = round(proportion * 100, 1)
    ) %>%
    select(`COG category` = cog_label, Direction = direction,
           `n genes` = n_genes, `proportion (%)`) %>%
    arrange(desc(`proportion (%)`))
}
```

```{r gwas-summary-table}
summarise_gwas <- function(label) {
  meta <- pyseer_files %>% filter(label == !!label)
  df   <- pyseer_data[[label]]
  egg  <- eggnog_lookup[[label]]

  df_annot <- df %>%
    left_join(egg, by = "variant") %>%
    mutate(
      significant     = p_fdr < 0.05,
      has_eggnog_name = !is.na(preferred) &
        !str_detect(preferred, regex("hypothetical", ignore_case = TRUE)),
      has_cog         = !is.na(cog) & cog != "-" & str_sub(cog, 1, 1) != "S"
    )

  sig <- df_annot %>% filter(significant)

  tibble(
    Analysis      = meta$title,
    Label         = label,
    Covariate     = ifelse(str_detect(label, "_covar$"), "Species-adjusted", "None"),
    n_tested      = nrow(df),
    n_significant = nrow(sig),
    n_human_associated     = sum(sig$beta > 0),
    n_non_human_associated = sum(sig$beta < 0),
    n_eggnog_annotated            = sum(sig$has_eggnog_name),
    n_eggnog_human_associated     = sum(sig$has_eggnog_name & sig$beta > 0),
    n_eggnog_non_human_associated = sum(sig$has_eggnog_name & sig$beta < 0),
    n_cog_annotated            = sum(sig$has_cog),
    n_cog_human_associated     = sum(sig$has_cog & sig$beta > 0),
    n_cog_non_human_associated = sum(sig$has_cog & sig$beta < 0)
  )
}

gwas_summary <- bind_rows(lapply(pyseer_files$label, summarise_gwas))
gwas_summary

gwas_summary_pretty <- gwas_summary %>%
  select(-Label) %>%
  rename(
    `Variants tested`             = n_tested,
    `Significant hits`            = n_significant,
    `Human-associated`            = n_human_associated,
    `Non-human-associated`        = n_non_human_associated,
    `eggNOG-annotated`            = n_eggnog_annotated,
    `eggNOG human-associated`     = n_eggnog_human_associated,
    `eggNOG non-human-associated` = n_eggnog_non_human_associated,
    `COG-annotated`               = n_cog_annotated,
    `COG human-associated`        = n_cog_human_associated,
    `COG non-human-associated`    = n_cog_non_human_associated
  )
```

# Save all GWAS tables to a single file
```{r gwas-write-xlsx}

top20_sheets <- setNames(
  lapply(pyseer_files$label, make_top20_eggnog_table),
  pyseer_files$label
)

cog_sheets <- list(
  cog_human_covar = make_cog_directional_table("human_covar"),
  cog_sp002404795         = make_cog_directional_table("sp002404795_human")
)

sheets <- c(list(gwas_summary = gwas_summary_pretty), top20_sheets, cog_sheets)
sheets <- sheets[sapply(sheets, function(x) nrow(x) > 0)]

write_xlsx(sheets, path = "gwas_tables.xlsx")
cat("Wrote", length(sheets), "sheets to gwas_tables.xlsx\n")
```
